---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "70%",
  warning = FALSE
)
```

# cforecast

**cforecast** is an R package for conditional forecasting and scenario analysis in vector autoregressive (VAR) models. It implements a Kalman filter–based framework for generating forecasts conditional on user-specified future paths of observable variables and provides tools for decomposing and interpreting scenario-driven forecast revisions.

The example below replicates an empirical experiment from:

> Caspi, I., & Ginker, T. (2026). *Conditional Forecasting with VARs: Dynamic Influence, Variable Importance, and Forecast Interpretation*.  
> https://doi.org/10.13140/RG.2.2.25225.51040

The illustration demonstrates the workflow for scenario design, conditional forecasting, and forecast attribution.

---

## Installation

Install the development version from GitHub:

```r
# install.packages("devtools")
devtools::install_github("timginker/cforecast")
```


## Empirical Illustration: Conditional Inflation Forecasting under Financial and Energy Price Scenarios

This section demonstrates how the conditional forecasting tools in `cforecast` can be used to:

1. Assess *ex ante* which conditioning assumptions are likely to materially affect a target forecast.
2. Attribute *ex post* forecast revisions to specific elements of a multi-period conditioning path.

We consider a stylized policy scenario combining:

- A tightening in financial conditions (corporate credit spreads), and  
- A temporary oil price disruption.

Such scenarios are typical in monetary policy and macro-financial stress-testing applications.

---

We use U.S. quarterly macroeconomic data (1986Q2–2015Q4) from FRED.  
The dataset includes the following series:

- **GDPC1** — Real GDP  
- **PCEPILFE** — Core PCE price index  
- **FEDFUNDS** — Federal funds rate  
- **BAA10YM** — Corporate credit spread (Moody’s Baa minus 10-year Treasury)  
- **DCOILWTICO** — WTI crude oil price  

The VAR includes five variables:

1. Real GDP growth  
2. Core PCE inflation  
3. Federal funds rate  
4. Corporate credit spread  
5. Oil price  

Below, we estimate a reduced-form VAR(2) with a constant term and compute a 20-quarter baseline (unconditional) forecast.


---


```{r message=FALSE}
suppressPackageStartupMessages({
library(cforecast)
library(vars)
library(dplyr)
library(lubridate)
library(ggplot2)
library(scales)
library(patchwork)
})

# Load packaged dataset
data(fred_macro)
data("DCOILWTICO_level")

# Restrict estimation sample
df <- fred_macro %>%
  filter(year(date) >= 1986,
         year(date) <= 2015)

# Estimate VAR(2)
fit <- VAR(df[, -1], p = 2, type = "const")

# Baseline forecast
pred_base <- predict(fit, n.ahead = 20)
```


---

## Scenario Design

We construct a two-variable conditioning path affecting:

- The corporate credit spread (BAA10YM)  
- The oil price (DCOILWTICO)

The scenario is designed as follows:

- **Credit spreads** increase by 200 basis points for three quarters and then gradually decay back toward baseline.
- **Oil prices** initially surge and then revert gradually.

This structure mimics a temporary macro-financial stress episode with persistent but fading effects.

The following plot creates the scenario and plots it:


```{r message=FALSE}
# Slicing Oil data
DCOILWTICO_level %>% 
  mutate(date = as.Date(date)) %>% 
  dplyr::filter(year(date) >= 1986, year(date) <= 2015) %>% 
  slice(-1) -> DCOILWTICO_level

# Baseline path for oil prices (reconstructed in levels)
pred_base <- predict(fit, n.ahead = 20)
oil_base <- rep(NA, 20)
oil_base[1] <- tail(DCOILWTICO_level$DCOILWTICO, 1) *
  (1 + pred_base$fcst$DCOILWTICO[1, 1] / 100)

for (i in 2:20) {
  oil_base[i] <- oil_base[i - 1] *
    (1 + pred_base$fcst$DCOILWTICO[i, 1] / 100)
}

# Imposed scenario paths for credit spreads and oil prices
delta <- 0.7
BAA10YM_future <- c(
  rep(tail(df$BAA10YM, 1) + 2, 3),
  (tail(df$BAA10YM, 1) + 2 * delta^(1:17))
)

DCOILWTICO_future <- c(
  tail(DCOILWTICO_level$DCOILWTICO, 1) * (1.155061^(1:4)),
  tail(DCOILWTICO_level$DCOILWTICO, 1) + 33.06 * delta^(1:16)
)

# Extended date vector covering the forecast horizon
dates_0 <- fred_macro$date[1:(nrow(df) + 20)]

# Construct plotting data for the credit-spread scenario input:
#   - y: historical series (in-sample)
#   - y_base: baseline (unconditional) VAR forecast appended after the sample end
#   - y_fcst: imposed scenario path appended after the sample end
# Missing values are used to prevent ggplot from drawing lines outside the
# intended segments (history vs forecast).

df_plot <- data.frame(
  date = dates_0,
  y = c(df$BAA10YM, rep(NA,length(BAA10YM_future))),
  y_base = c(rep(NA,nrow(df)-1),
             tail(df$BAA10YM,1),pred_base$fcst$BAA10YM[,1]),
  y_fcst = c(rep(NA,nrow(df)-1),
             tail(df$BAA10YM,1),BAA10YM_future)
)

# Plot the credit-spread path:
# Solid line: historical data
# Dashed line: imposed scenario path used for conditioning
# Dotted line: baseline path implied by the VAR
p_baa <- ggplot(df_plot, aes(x = date)) +
  geom_line(aes(y = y), color = "black", linewidth = 0.8) +
  geom_line(
    aes(y = y_fcst),
    color = "black",
    linewidth = 0.7,
    linetype = "dashed"
  ) +
  geom_line(
    aes(y = y_base),
    color = "black",
    linewidth = 0.7,
    linetype = "dotted"
  )+
  labs(
    title = "Moody's Seasoned Baa Corporate Bond Yield Relative to \nYield on 10-Year Treasury Constant Maturity",
    #subtitle = "History (solid) and scenario (dashed)",
    x = NULL,
    y = "Percentage points"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 12),
    axis.text.y = element_text(size = 8)
    #plot.title = element_text(face = "bold")
  )


# Construct plotting data for the oil price scenario input (in levels):
#   - y: historical oil price (levels series)
#   - y_fcst: imposed scenario oil price path in levels
#   - y_base: baseline oil price path reconstructed from VAR forecasts
# As above, NA padding separates the historical segment from forecast segments.

df_plot_oil <- data.frame(
  date = dates_0,
  y = c(DCOILWTICO_level$DCOILWTICO, rep(NA,length(DCOILWTICO_future))),
  y_fcst = c(rep(NA,nrow(df)-1), 
             tail(DCOILWTICO_level$DCOILWTICO,1),DCOILWTICO_future),
  y_base = c(rep(NA,nrow(df)-1), 
             tail(DCOILWTICO_level$DCOILWTICO,1),oil_base)
)

# Plot the oil price path:
# Solid line: historical data
# Dashed line: imposed scenario path used for conditioning
# Dotted line: baseline path implied by the VAR (in levels)
p_oil <- ggplot(df_plot_oil, aes(x = date)) +
  geom_line(aes(y = y), color = "black", linewidth = 0.8) +
  geom_line(
    aes(y = y_fcst),
    color = "black",
    linewidth = 0.7,
    linetype = "dashed"
  ) +
  geom_line(
    aes(y = y_base),
    color = "black",
    linewidth = 0.7,
    linetype = "dotted"
  )+
  labs(
    title = "WTI Crude Oil Price",
    #subtitle = "History (solid) and scenario (dashed)",
    x = NULL,
    y = "USD per barrel"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 12),
    axis.text.y = element_text(size = 8)
  )


# Align x-axis limits across the two panels to ensure consistent time coverage
# and facilitate direct visual comparison of the two conditioning paths.
x_rng <- range(c(df_plot_oil$date, df_plot$date), na.rm = TRUE)

# Combine the two scenario-input panels into a single figure:
# Oil price on top; credit spread below. X-axis annotations are removed from
# the top panel to avoid duplication and improve readability.
combined_plot_scenarios <-
  (p_oil + scale_x_date(limits = x_rng) +
     theme(axis.title.x = element_blank(),
           axis.text.x  = element_blank(),
           axis.ticks.x = element_blank())) /
  (p_baa + scale_x_date(limits = x_rng))


suppressWarnings(combined_plot_scenarios)
```



The object `cond_path` contains the multi-period conditioning trajectory supplied to the conditional forecasting procedure.


---

## Interpreting Variable Importance

We begin by examining overall variable importance in the conditional forecast of core inflation. These measures are computed *ex ante*, given the conditioning design: the future paths of the corporate credit spread (Moody’s Baa–10Y Treasury spread) and WTI crude oil prices are constrained, while the remaining variables are left unconstrained.

The decomposition therefore quantifies the relative contribution of:

1. The imposed conditioning paths, and  
2. Historical information embedded in the data.

Importantly, this assessment does not require realized future values.

The plot below reports the overall variable importance:

```{r}
v_imp=variable_importance_stat(fit=fit,
                               cond_var = 4:5,
                               target_var = 2,
                               horizon = 20)

# Visualize variable importance as stacked shares by horizon (normalized to 100%)
plt = ggplot(v_imp$variable_importance,
             aes(x = factor(horizon),
                 y = share,
                 fill = variable)) +
  geom_col(position = "fill", width = 0.75, color = "white", linewidth = 0.2) +
  scale_y_continuous(
    labels = percent_format(accuracy = 1),
    expand = expansion(mult = c(0, 0.02))
  ) +
  scale_fill_viridis_d(option = "cividis", end = 0.9) +
  labs(
    title = "Decomposition of Conditional Core Inflation Forecast",
    subtitle = "Overall variable importance by forecast horizon",
    x = "Forecast horizon (quarters ahead)",
    y = "Share of information (percent)",
    fill = NULL
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 11),
    plot.subtitle = element_text(size = 10),
    axis.title = element_text(size = 11),
    axis.text  = element_text(size = 10),
    legend.position = "top",
    legend.text = element_text(size = 10),
    legend.key.size = unit(0.35, "cm"),
    axis.line = element_line(linewidth = 0.3),
    axis.ticks = element_line(linewidth = 0.3)
  )

plt 
```


The results indicate that the importance of the credit spread increases steadily with the forecast horizon. This pattern suggests that financial conditions—particularly corporate credit risk—play an increasingly prominent role in shaping medium-term inflation dynamics. This finding is consistent with evidence that credit spreads contain predictive information for demand-side pressures (e.g., Lopez-Salido et al., 2017; Caldara and Herbst, 2019).

By contrast, oil prices contribute more modestly to the core inflation forecast. Their importance rises slightly at longer horizons—an intuitive result given that core inflation excludes energy components and oil affects inflation only indirectly through production costs and aggregate demand.

The federal funds rate, although unconstrained in the scenario, contributes through its historical values. Its importance peaks at a lag of roughly five quarters, consistent with standard estimates of monetary policy transmission (e.g., Christiano et al., 1996).

Finally, GDP growth receives minimal weight in the decomposition. This reflects its weak marginal signal once financial and commodity variables are accounted for and aligns with evidence that real activity indicators offer limited incremental predictive content for inflation (Stock and Watson, 2007).

---

## Marginal Variable Importance

The plot below reports the marginal importance:

```{r}

plt_mimp = ggplot(subset(v_imp$marginal_variable_importance,share!=0),
             aes(x = factor(horizon),
                 y = share,
                 fill = variable)) +
  geom_col(position = "fill", width = 0.75, color = "white", linewidth = 0.2) +
  scale_y_continuous(
    labels = percent_format(accuracy = 1),
    expand = expansion(mult = c(0, 0.02))
  ) +
  scale_fill_viridis_d(option = "cividis", end = 0.9) +
  labs(
    title = "Decomposition of Conditional Core Inflation Forecast",
    subtitle = "Marginal variable importance by forecast horizon",
    x = "Forecast horizon (quarters ahead)",
    y = "Share of information (percent)",
    fill = NULL
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 11),
    plot.subtitle = element_text(size = 10),
    axis.title = element_text(size = 11),
    axis.text  = element_text(size = 10),
    legend.position = "top",
    legend.text = element_text(size = 10),
    legend.key.size = unit(0.35, "cm"),
    axis.line = element_line(linewidth = 0.3),
    axis.ticks = element_line(linewidth = 0.3)
  )


plt_mimp
```


Marginal variable importance isolates the contribution of the imposed future constraints. The decomposition is dominated by the corporate credit spread, while oil prices gain importance gradually at longer horizons due to their indirect and lagged transmission to core inflation.

Taken together, the overall and marginal importance measures highlight an important distinction:

- **Overall importance** reflects both historical information and imposed paths.  
- **Marginal importance** isolates the incremental role of future constraints.

The results emphasize the dominant role of the credit spread in shaping the conditional inflation forecast and the delayed, indirect influence of oil prices.


---


# Disclaimer

The views expressed here are solely of the author and do not necessarily represent the views of the Bank of Israel. 

Please note that `cforecast` is still under development and may contain bugs or other issues that have not yet been resolved. While we have made every effort to ensure that the package is functional and reliable, we cannot guarantee its performance in all situations. 

We strongly advise that you regularly check for updates and install any new versions that become available, as these may contain important bug fixes and other improvements. By using this package, you acknowledge and accept that it is provided on an "as is" basis, and that we make no warranties or representations regarding its suitability for your specific needs or purposes.
